Component response rate variation underlies the stability of complex systems

### 2020-03-02 notebook/ms.Rmd
Reworking the model so that simulations are introduced in a more seemless way

Equation \ref{defM} thereby serves as a null model to investigate how variation in component response rate ($\sigma_{\gamma}$) affects complex systems. In the absence of such variation ($\sigma_{\gamma} = 0$), $\gamma$ is set to the identity matrix (diagonal elements all equal 1) and $\textbf{M} = \textbf{A}$.

@@

The increase caused by $\sigma^{2}_\gamma$ can be visualised from the eigenvalue spectra of $\textbf{A}$ versus $\textbf{M} = \gamma\textbf{A}$ (Figure 1). Given $d = 0$ and $C = 1$, the distribution of eigenvalues of $\textbf{A}$ and $\textbf{M}$ lie within a circle of a radius $\sigma^{2}_{A}\sqrt{S}$ and $\sigma^{2}_{M}\sqrt{S}$, respectively (Figure 1a vs. 1b). When components are self-regulating (e.g., $d = -1$), positive $\sigma^{2}_\gamma$ changes the distribution of eigenvalues, potentially affecting stability (Figure 1c vs. 1d).

Given $\sigma^{2}_\gamma = 0$, $\max(\Re(\lambda))$ decrease linearly with $\rho$ such that,

@@

When $\rho$ values are negative, such as when $\textbf{M}$ models a predator-prey system in which $M_{ij}$ and $M_{ji}$ have opposing signs, stability is expected to increase [@Allesina2012]. When elements of $\gamma$ vary independently, the magnitude of $\rho$ is decreased because $\sigma^{2}_{\gamma}$ increases the variance of $M_{ij}$ without affecting the covariance between $M_{ij}$ and $M_{ji}$,

@@

Because $\gamma$ does not affect $cov(M_{ij}, M_{ji})$, the sign of $\rho$ remains unchanged.

**Numerical simulations**. I used numerical simulations to isolate the effect of $\sigma_{\gamma}$ on the stability of $\textbf{M} = \gamma \textbf{A}$ by comparing $\sigma_{\gamma} > 0$ versus $\sigma_{\gamma} = 0$. Values of $\gamma_{ii}$ were sampled such that $E[d]$ and $E[\mu]$ also remained unchanged (in practice, diagonal elements of $\textbf{M}$ were standardised so that mean values were identical before and after adding $\gamma$).

@@

>>>>

Equation \ref{defM} thereby serves as a null model to investigate how variation in component response rate ($\sigma_{\gamma}$) affects complex systems. In the absence of such variation ($\sigma_{\gamma} = 0$), $\gamma$ is set to the identity matrix (diagonal elements all equal 1) and $\textbf{M} = \textbf{A}$. Under these conditions, eigenvalues of $\textbf{M}$ are distributed uniformly [@Tao2010] in a circle centred at $(-d, 0)$ with a radius of $\sigma_{M} \sqrt{SC}$ [@May1972] (Figure 1a).

@@

The increase caused by $\sigma^{2}_\gamma$ can be visualised from the eigenvalue spectra of $\textbf{A}$ versus $\textbf{M} = \gamma\textbf{A}$ (Figure 1). Given $d = 0$ and $C = 1$, the distribution of eigenvalues of $\textbf{A}$ and $\textbf{M}$ lie within a circle of a radius $\sigma^{2}_{A}\sqrt{S}$ and $\sigma^{2}_{M}\sqrt{S}$, respectively (Figure 1a vs. 1b). If $d \neq 0$, positive $\sigma^{2}_\gamma$ changes the distribution of eigenvalues [@Ahmadian2015; @Gibbs2017; @Stone2017], potentially affecting stability (Figure 1c vs. 1d).

Given $\sigma^{2}_\gamma = 0$, $\max(\Re(\lambda))$ decrease linearly with $\rho$ such that [@Tang2014c],

@@

If $\rho < 0$, such as when $\textbf{M}$ models a predator-prey system in which $M_{ij}$ and $M_{ji}$ have opposing signs, stability increases [@Allesina2012]. If diagonal elements of $\gamma$ vary independently, the magnitude of $\rho$ is decreased because $\sigma^{2}_{\gamma}$ increases the variance of $M_{ij}$ without affecting the expected covariance between $M_{ij}$ and $M_{ji}$,

@@

Neverthless, numerical simulations reveal that $\sigma^{2}_{\gamma} > 0$ results in a nonlinear relationship between $\rho$ and $\max(\Re(\lambda))$, which can sometimes increase the stability of the system.

**Numerical simulations**. I used numerical simulations to isolate the effect of $\sigma_{\gamma}$ on system stability by comparing the same $\textbf{M} = \gamma \textbf{A}$ given $\sigma_{\gamma} > 0$ versus $\sigma_{\gamma} = 0$. Values of $\gamma_{ii}$ were sampled from a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). For $\sigma_{\gamma} = 0$, diagonal elements were standardised to the mean value of sampled $\gamma_{ii}$ so that $E[d]$ of $\textbf{M}$ were identical.

@@

<!---  (in practice, diagonal elements of $\textbf{M}$ were standardised so that mean values were identical before and after adding $\gamma$) --->
